---
title: "Analysis of Spanish and English Vowel Space Areas"
author: "Annie Helms (annie_helms@berkeley.edu)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "cairo_pdf")
#options(scipen = 999) # avoid scientific notation
```

Library packages

```{r message = FALSE, warning = FALSE}
library(lmerTest) #mixed effects linear regression
library(dplyr)
library(kableExtra) # formats dataframes in html output
library(emmeans) # Tukey tests
library(ggplot2)
```

Set working directory and import data set of areas by speaker in the Late Show corpus.

```{r message = FALSE, echo = FALSE}
setwd("/Users/atarv/Box Sync/PhD/github/vsa_langdom/")
```

```{r}
spa = read.csv("data/spa_ratios.csv", 
                  stringsAsFactors = TRUE)
kbl(head(spa), table.attr = "style = \"color: black;\"") %>% # piped to allow for kable formatting
  kable_styling(bootstrap_options = c("striped", # alternates row color in table
                                      "hover", # rows are highlighted when hovered over
                                      "condensed", # row height is condensed
                                      "responsive")) %>% # horizontal scrolling across table
  scroll_box(width = "100%")
```

```{r}
dim(spa)
```

Create subsets of data for each cutoff

```{r}
cut_10 = spa %>%
  filter(Cutoff==0.1)
cut_15 = spa %>%
  filter(Cutoff==0.15)
cut_20 = spa %>%
  filter(Cutoff==0.2)
cut_25 = spa %>%
  filter(Cutoff==0.25)
cut_30 = spa %>%
  filter(Cutoff==0.3)
```

Set up mixed effects linear regression, with DV of area, IV of stress, and random intercept of speaker. Run for each cutoff level.

```{r}
# submit to mixed effects linear regression model
model_10 = glm(Area_ratio ~ Dom + avg_dur, data = cut_10)
summary(model_10)
```

According to the model output above, neither avg_dur, stress, or language dominance is a significant predictor of vowel space area.

```{r}
# submit to mixed effects linear regression model
model_15 = glm(Area_ratio ~ Dom + avg_dur, data = cut_15)
summary(model_15)
```
According to the model output above, neither avg_dur, stress, or language dominance is a significant predictor of vowel space area.

```{r}
# submit to mixed effects linear regression model
model_20 = glm(Area_ratio ~ Dom + avg_dur, data = cut_20)
summary(model_20)
```

According to the model output above, neither avg_dur, stress, or language dominance is a significant predictor of vowel space area.

```{r}
# submit to mixed effects linear regression model
model_25 = glm(Area_ratio ~ Dom + avg_dur, data = cut_25)
summary(model_25)
```

Post-hoc:

```{r}
emmeans(model_25, list(pairwise ~ is_stress * Dom), adjust="tukey")
```

According to the model output above, neither avg_dur, stress, or language dominance is a significant predictor of vowel space area.

```{r}
# submit to mixed effects linear regression model
model_30 = glm(Area_ratio ~ Dom + avg_dur, data = cut_30)
summary(model_30)
```

According to the model output above, neither avg_dur, stress, or language dominance is a significant predictor of vowel space area.


# Now just CBAS speakers

```{r}
cbas = read.csv("data/cbas_ratios.csv", 
                  stringsAsFactors = TRUE)
kbl(head(cbas), table.attr = "style = \"color: black;\"") %>% # piped to allow for kable formatting
  kable_styling(bootstrap_options = c("striped", # alternates row color in table
                                      "hover", # rows are highlighted when hovered over
                                      "condensed", # row height is condensed
                                      "responsive")) %>% # horizontal scrolling across table
  scroll_box(width = "100%")
```

```{r}
dim(cbas)
```

Create subsets of data for each cutoff

```{r}
cut_10 = cbas %>%
  filter(Cutoff==0.1)
cut_15 = cbas %>%
  filter(Cutoff==0.15)
cut_20 = cbas %>%
  filter(Cutoff==0.2)
cut_25 = cbas %>%
  filter(Cutoff==0.25)
cut_30 = cbas %>%
  filter(Cutoff==0.3)
```

Set up mixed effects linear regression, with DV of area, IV of stress, and random intercept of speaker. Run for each cutoff level.

```{r}
# submit to mixed effects linear regression model
model_10 = lmer(Area_ratio ~ Dom * Language + avg_dur * Language + (1|Participant), 
                data = cut_10, REML = FALSE)
summary(model_10)
```

According to the model output above, neither avg_dur, stress, or language dominance is a significant predictor of vowel space area.

Post-hoc:

```{r}
emmeans(model_10, list(pairwise ~ is_stress * Language), adjust="tukey")
```

Stress is only a predictor of English vowel space area (i.e. not Spanish) and stressed English is significantly larger than stressed Spanish.

```{r}
# submit to mixed effects linear regression model
model_15 = lmer(Area_ratio ~ Dom * Language + avg_dur * Language + (1|Participant), 
                data = cut_15, REML = FALSE)
summary(model_15)
```
According to the model output above, neither avg_dur, stress, or language dominance is a significant predictor of vowel space area.

Post-hoc:

```{r}
emmeans(model_15, list(pairwise ~ is_stress * Language), adjust="tukey")
```
Stressed English is bigger than stressed Spanish. 

```{r}
# submit to mixed effects linear regression model
model_20 = lmer(Area_ratio ~ Dom * Language + avg_dur * Language + (1|Participant), 
                data = cut_20, REML = FALSE)
summary(model_20)
```

According to the model output above, neither avg_dur, stress, or language dominance is a significant predictor of vowel space area.

Post-hoc:

```{r}
emmeans(model_20, list(pairwise ~ is_stress * Language), adjust="tukey")
```
Stressed English bigger than stressed Spanish.



```{r}
# submit to mixed effects linear regression model
model_25 = lmer(Area_ratio ~ Dom * Language + avg_dur * Language + (1|Participant), 
                data = cut_25, REML = FALSE)
summary(model_25)
```

Post-hoc:

```{r}
emmeans(model_25, list(pairwise ~ Language * Dom), adjust="tukey")
```
What? Stress is a significant predictor of vowel space area for early bilinguals, regardless of language. However, the same is not true for L2-Spanish speakers. 

```{r}
emmeans(model_25, list(pairwise ~ is_stress * Language), adjust="tukey")
```



```{r}
# submit to mixed effects linear regression model
model_30 = lmer(Area_ratio ~ Dom * Language + avg_dur * Language + (1|Participant), 
                data = cut_30, REML = FALSE)
summary(model_30)
```

According to the model output above, neither avg_dur, stress, or language dominance is a significant predictor of vowel space area.

Post-hoc:

```{r}
emmeans(model_30, list(pairwise ~ is_stress * Dom), adjust="tukey")
```

Post-hoc:

```{r}
emmeans(model_30, list(pairwise ~ is_stress * Language), adjust="tukey")
```

Now size difference of stressed productions between English and Spanish disappears

Plot size across cutoff, language dominance, and language

```{r}
spa_area_ratio_plot = ggplot(data = spa, aes(x = Dom, y = Area_ratio, color = Dom)) + 
  geom_boxplot()
spa_area_ratio_plot
```

```{r}
cbas_area_ratio_plot = ggplot(data = cbas, aes(x = Language:Dom, y = Area_ratio, color = Language:Dom)) + 
  geom_boxplot()
cbas_area_ratio_plot
```
Look at variability across speakers/homogeneity.


# Now just Spanish, but divide speakers in Eng exposure (True, False)

```{r}
spa_cbas = read.csv("data/spa_cbas.csv", 
                  stringsAsFactors = TRUE)
kbl(head(spa_cbas), table.attr = "style = \"color: black;\"") %>% # piped to allow for kable formatting
  kable_styling(bootstrap_options = c("striped", # alternates row color in table
                                      "hover", # rows are highlighted when hovered over
                                      "condensed", # row height is condensed
                                      "responsive")) %>% # horizontal scrolling across table
  scroll_box(width = "100%")
```

```{r}
dim(spa_cbas)
```

Create subsets of data for each cutoff

```{r}
cut_10 = spa_cbas %>%
  filter(Cutoff==0.1)
cut_15 = spa_cbas %>%
  filter(Cutoff==0.15)
cut_20 = spa_cbas %>%
  filter(Cutoff==0.2)
cut_25 = spa_cbas %>%
  filter(Cutoff==0.25)
cut_30 = spa_cbas %>%
  filter(Cutoff==0.3)
```

Set up mixed effects linear regression, with DV of area, IV of stress, and random intercept of speaker. Run for each cutoff level.

```{r}
# submit to mixed effects linear regression model
model_10 = lmer(Area ~ is_stress * Corpus + avg_dur + (1|Participant),
                   data = cut_10, REML = FALSE)
summary(model_10)
```

According to the model output above, neither avg_dur, stress, or language dominance is a significant predictor of vowel space area.

```{r}
# submit to mixed effects linear regression model
model_15 = lmer(Area ~ is_stress * Corpus + avg_dur + (1|Participant),
                   data = cut_15, REML = FALSE)
summary(model_15)
```

According to the model output above, neither avg_dur, stress, or language dominance is a significant predictor of vowel space area.

```{r}
# submit to mixed effects linear regression model
model_20 = lmer(Area ~ is_stress * Corpus + avg_dur + (1|Participant),
                   data = cut_20, REML = FALSE)
summary(model_20)
```

According to the model output above, neither avg_dur, stress, or language dominance is a significant predictor of vowel space area.

```{r}
# submit to mixed effects linear regression model
model_25 = lmer(Area ~ is_stress * Corpus + avg_dur + (1|Participant),
                   data = cut_25, REML = FALSE)
summary(model_25)
```

```{r}
# submit to mixed effects linear regression model
model_30 = lmer(Area ~ is_stress * Corpus + avg_dur + (1|Participant),
                   data = cut_30, REML = FALSE)
summary(model_30)
```


